---
title: "Heart Failure Research"
author: "Gianni Conde, Joshua Biggs-Bauer, Joaquin Rodarte"
date: "Winter 2022"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal
To determine the event of death due to heart failure based on clinical measurements utilizing supervised machine learning and data mining methods.


# About the Data
The data was retrieved from Kaggle and was originally collected from April to December of 2015 by the Allied Hospital in Faisalabad and the Faisalabad Institute of Cardiology, both residing in Punjab, Pakistan. The data contains patient medical information regarding heart failure with 299 patients (observations) and 13 clinical measurements (variables). These clinical measurements included age, anemia, creatinine phosphokinase (CPK), diabetes, ejection fraction, high blood pressure, platelets, serum creatinine, serum sodium, sex, smoking, time, and death event.

Link(s): https://www.kaggle.com/datasets/andrewmvd/heart-failure-clinical-data


# Role
My role in this analysis was to perform data cleansing and preparation, assist in creating exploratory visualizations, and to develop association rule mining models.


# Libaries
```{r}
library(arules)
library(arulesViz)
library(caret)
library(corrplot)
library(DescTools)
library(directlabels)
library(dplyr)
library(e1071)
library(gbm)
library(ggplot2)
library(ggfortify)
library(ggthemes)
library(hrbrthemes)
library(ISLR)
library(kableExtra)
library(lightgbm)
library(mlr)
library(patchwork)
library(pROC)
library(randomForest)
library(randomForestExplainer)
library(ranger)
library(RColorBrewer)
library(reshape2)
library(rminer)
library(rpart)
library(rpart.plot)
library(rsample)
library(scales)
library(sjPlot)
library(skimr)
library(stats)
library(survminer)
library(survival)
library(tidyverse)
library(tree)
library(vip)
library(viridis)
```

# Loading the Data
```{r}
heartFailure <- read.csv("C:/Users/cassh/OneDrive/Desktop/Data Mining/Project/heart_failure_clinical_records_dataset.csv")
```

## Initial Inspecting of the Data 
```{r}
dim(heartFailure)
view(heartFailure)
```
299 patients (observations), 13 clinical measurements (variables).


# Data Cleansing and Preparation
## Checking for missing values
```{r}
sum(is.na(heartFailure))
```
0 missing values.


## Removing 'time' column and adding patient 'id' index
```{r}
heartFailure$time <- NULL
heartFailure <- tibble::rowid_to_column(heartFailure, "id")
str(heartFailure)
view(heartFailure)
```
After removing and adding the two variables, the remaining columns are either numeric or integers.


# Exploratory Data Analysis
## Observing numerical data
```{r}
numData <- heartFailure %>% select(1:2,4,6,8:10)
str(numData)
dim(numData) # 299 patients, 7 clinical measures
colnames(numData)
view(numData)
summary(numData)
view(summary(numData))
```
The newly created data frame features quantifiable measurements, except for 'id'. This data frame includes 299 patients and 7 clinical measures.


## Observing binary data
```{r}
binaryData <- (heartFailure %>% select(3,5,7,11:13))
str(binaryData)
dim(binaryData)
colnames(binaryData)
view(binaryData)
```
The newly created data frame features qualitative measurements including 299 patients and 6 clinical measures. 


```{r}
table(heartFailure$anaemia)
table(heartFailure$diabetes)
table(heartFailure$high_blood_pressure)
table(heartFailure$sex)
table(heartFailure$smoking)
table(heartFailure$DEATH_EVENT)
```
According to the tables, 129 patients have anemia, while 170 do not. Regarding diabetes, 174 patients were positively diagnosed, while 125 were not. Due to the similar ratios, it is reasonable to assume that anemia is strongly correlated to diabetes. 

105 patients had high blood pressure, while 194 did not. 105 of the patients were female, while 194 were male. Despite their identical ratios, it would not be wise to assume that they are correlated since it would be unlikely that all patients with high blood pressure were female. 

96 patients were smokers, while 203 were non-smokers. Regarding death event, 96 patients died while 203 survived. Based on the scope of this analysis, it is somewhat reasonable to act under the assumption that smoking is strongly correlated to death by heart failure. However, further analysis is required.


## Plotting Distributions, Correlations, and Box Plots
### Anemia vs. Death Event
```{r}
ggplot(heartFailure, aes(x = factor(anaemia), fill = factor(DEATH_EVENT))) +
  geom_bar(stat = "count", position = "stack", show.legend = FALSE) +
  scale_x_discrete(labels  = c("0 (not anemic)", "1 (anemic)")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Anaemia") +
  labs(title = "Distribution of Anaemia against DEATH_EVENT") +
  geom_label(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5),
             size = 5)
```
### Diabetes vs. Death Event
```{r}
ggplot(heartFailure, aes(x = factor(diabetes), fill = factor(DEATH_EVENT))) +
  geom_bar(stat = "count", position = "stack", show.legend = FALSE) +
  scale_x_discrete(labels  = c("0 (not diabetic)", "1 (diabetic)")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Diabetes") +
  labs(title = "Distribution of Diabetes against DEATH_EVENT") +
  geom_label(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5),
             size = 5)
```
### High Blood Pressure vs. Death Event
```{r}
ggplot(heartFailure, aes(x = factor(high_blood_pressure), fill = factor(DEATH_EVENT))) +
  geom_bar(stat = "count", position = "stack", show.legend = FALSE) +
  scale_x_discrete(labels  = c("0 (stable)", "1 (high)")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "High Blood Pressure") +
  labs(title = "Distribution of High Blood Pressure against DEATH_EVENT") +
  geom_label(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5),
             size = 5)
```
### Sex vs. Death Event
```{r}
ggplot(heartFailure, aes(x = factor(sex), fill = factor(DEATH_EVENT))) +
  geom_bar(stat = "count", position = "stack", show.legend = FALSE) +
  scale_x_discrete(labels  = c("0 (Female)", "1 (Male)")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Sex") +
  labs(title = "Distribution of Sex against DEATH_EVENT") +
  geom_label(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5),
             size = 5)
```
### Smoking vs. Death Event
```{r}
ggplot(heartFailure, aes(x = factor(smoking), fill = factor(DEATH_EVENT))) +
  geom_bar(stat = "count", position = "stack", show.legend = FALSE) +
  scale_x_discrete(labels  = c("0 (non-smoker)", "1 (smoker)")) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Smoking") +
  labs(title = "Distribution of Smoking against DEATH_EVENT") +
  geom_label(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5),
             size = 5)
```


### Distribtuion of Age and Death Event
```{r}
Viz1 <-ggplot(heartFailure, aes(x=age)) + 
  geom_histogram(binwidth = 5, colour="black", fill="lightblue", alpha=0.8) +
  geom_density(eval(bquote(aes(y=..count..*5))), colour="purple", fill="purple", 
  alpha=0.3) + scale_x_continuous(breaks = seq(40,100,10)) +
  geom_vline(xintercept = 65, linetype="dashed") + 
  annotate("text", x=50, y=45, label="Age <65", size=2.5, color="purple") + 
  annotate("text", x=80, y=45, label="Age >= 65", size=2.5, color="dark red") + 
  labs(title="Age Distribution") + theme_minimal(base_size = 8)
Viz1
```
The histogram shaded in blue displays the layer of patient age and the purple overlapping portion of the graph represents the distribution of death events among patients belonging to their age groups. Since there were no participants under the age of 40 and over the age of 95, the overlapping distribution of death events among patient ages begin and end at those points.


### Serum Creatinine Distribution
```{r}
Viz2 <- ggplot(heartFailure, aes(x=serum_creatinine)) + 
  geom_histogram(binwidth = 5, colour="black", fill="lightblue", alpha=0.8) +
  geom_density(eval(bquote(aes(y=..count..*1))), colour="purple", fill="purple", 
  alpha=0.3) + scale_x_continuous(breaks = seq(0,2,0.1)) +
  labs(title="Serum Creatinine Distribution") + theme_minimal(base_size = 8)
Viz2
```
### Correlation Plot
```{r}
C = cor(heartFailure)
corrplot(C, method = 'square', order = 'FPC',type = 'lower', 
         diag = FALSE, main = 'Correlation Plot')
```
The correlation plots shows that smoking and sex had the highest positive correlation, followed by serum creatinine and death event, and age and death event. The most negatively correlated data point not including 'id' was ejection fraction and death event, implying that a patient’s ejection fraction has little effect on their mortality rate.


### Age and Death Event Box Plots
```{r}
hf_death <- filter(heartFailure,DEATH_EVENT==1)
hf_noDeath <- filter(heartFailure,DEATH_EVENT==0)

boxplot(hf_death$age,ylab="Age of the Deceased Patients" , xlab="Patients - Death event", col="red")
boxplot(hf_noDeath$age,ylab="Age of Live Patients" , xlab="Patients - Live", col="blue")$out
summary(hf_death)
summary(hf_noDeath)
```

# Model Preparation
## 1st Data Split
```{r}
set.seed=9
trainTestSplit1 <- sample(2, nrow(heartFailure), replace=TRUE, prob=c(0.8,0.2))
train1 = heartFailure[trainTestSplit1==1,]
test1 = heartFailure[trainTestSplit1==2,]
dim(train1)
dim(test1)
```

## 2nd Data Split
```{r}
set.seed=20
trainTestSplit2 <- sample(2, nrow(heartFailure), replace=TRUE, prob=c(0.7,0.3))
train2 = heartFailure[trainTestSplit2==1,]
test2 = heartFailure[trainTestSplit2==2,]
dim(train2)
dim(test2)
```

## Drawing Confusion Matrix
```{r}
## Function for drawing confusion matrix
draw_confusion_matrix <- function(cm) {
  total <- sum(cm$table)
  res <- as.numeric(cm$table)
  # Set up the colors (and gradients) RColorBrewer.
  greenPalette <- c("#F7FCF5","#E5F5E0","#C7E9C0","#A1D99B","#74C476","#41AB5D","#238B45","#006D2C","#00441B")
  redPalette <- c("#FFF5F0","#FEE0D2","#FCBBA1","#FC9272","#FB6A4A","#EF3B2C","#CB181D","#A50F15","#67000D")
  getColor <- function (greenOrRed = "green", amount = 0) {
    if (amount == 0)
      return("#FFFFFF")
    palette <- greenPalette
    if (greenOrRed == "red")
      palette <- redPalette
    colorRampPalette(palette)(100)[10 + ceiling(90 * amount / total)]
  }
  # CM Layout
  layout(matrix(c(1,1,2)))
  par(mar=c(2,2,2,2))
  plot(c(100, 345), c(300, 450), type = "n", xlab="", ylab="", xaxt='n', yaxt='n')
  title('CONFUSION MATRIX', cex.main=2)
  # The Matrix 
  classes = colnames(cm$table)
  rect(150, 430, 240, 370, col=getColor("green", res[1]))
  text(195, 435, classes[1], cex=1.2)
  rect(250, 430, 340, 370, col=getColor("red", res[3]))
  text(295, 435, classes[2], cex=1.2)
  text(125, 370, 'Predicted', cex=1.3, srt=90, font=2)
  text(245, 450, 'Actual', cex=1.3, font=2)
  rect(150, 305, 240, 365, col=getColor("red", res[2]))
  rect(250, 305, 340, 365, col=getColor("green", res[4]))
  text(140, 400, classes[1], cex=1.2, srt=90)
  text(140, 335, classes[2], cex=1.2, srt=90)
  # Show the results
  text(195, 400, res[1], cex=1.6, font=2, col='white')
  text(195, 335, res[2], cex=1.6, font=2, col='white')
  text(295, 400, res[3], cex=1.6, font=2, col='white')
  text(295, 335, res[4], cex=1.6, font=2, col='white')
  # specifics, specifics, and more specifics 
  plot(c(100, 0), c(100, 0), type = "n", xlab="", ylab="", main = "DETAILS", xaxt='n', yaxt='n')
  text(10, 85, names(cm$byClass[1]), cex=1.2, font=2)
  text(10, 70, round(as.numeric(cm$byClass[1]), 3), cex=1.2)
  text(30, 85, names(cm$byClass[2]), cex=1.2, font=2)
  text(30, 70, round(as.numeric(cm$byClass[2]), 3), cex=1.2)
  text(50, 85, names(cm$byClass[5]), cex=1.2, font=2)
  text(50, 70, round(as.numeric(cm$byClass[5]), 3), cex=1.2)
  text(70, 85, names(cm$byClass[6]), cex=1.2, font=2)
  text(70, 70, round(as.numeric(cm$byClass[6]), 3), cex=1.2)
  text(90, 85, names(cm$byClass[7]), cex=1.2, font=2)
  text(90, 70, round(as.numeric(cm$byClass[7]), 3), cex=1.2)
  # How accurate were we
  text(30, 35, names(cm$overall[1]), cex=1.5, font=2)
  text(30, 20, round(as.numeric(cm$overall[1]), 3), cex=1.4)
  text(70, 35, names(cm$overall[2]), cex=1.5, font=2)
  text(70, 20, round(as.numeric(cm$overall[2]), 3), cex=1.4)
  }
```


# Random Forest
## 1st Attempt
```{r}
rforest<- randomForest(factor(DEATH_EVENT) ~. , data=train1, ntree=1000, importance=TRUE)
summary(rforest)
imp <- varImp(rforest)
varImpPlot(rforest)
# Predict the 1st random forest
rpredict <- predict(rforest, test1, type="class")
cm1 <- confusionMatrix(rpredict, as.factor(test1$DEATH_EVENT))
draw_confusion_matrix(cm1)
```

## 2nd Attempt
```{r}
moreTrees <- randomForest(factor(DEATH_EVENT) ~. , data=train2, ntree=1000, importance=TRUE)
summary(moreTrees)
imp <- varImp(moreTrees)
varImpPlot(moreTrees)
# Predict the 2nd random forest
treePredict <- predict(moreTrees, test2, type="class")
cm2 <- confusionMatrix(treePredict, as.factor(test2$DEATH_EVENT))
draw_confusion_matrix(cm2)
```
The 1st Random Forest attempt yielded the best results in terms of every metric, essentially making it the best model of the two.


# Support Vector Machine (SVM)
## 1st Attempt
```{r}
svmmodel <- svm(formula=DEATH_EVENT~., data=train1, type="C-classification", kernel="linear")
# prediction
pred = predict(svmmodel, newdata=test2)
# Confusion matrix
cm3 <- confusionMatrix(pred, as.factor(test2$DEATH_EVENT))
draw_confusion_matrix(cm3)
```

## 2nd Attempt
```{r}
# tune SVM
tune <- tune.svm(x=train1[-3],y=train1[3], cost = seq( 0.2,0.3, 1))
tune$best.parameters          
# SVM model              
svmmodel2 <-svm(formula=DEATH_EVENT~., data=train1, type="nu-classification", kernel="linear", cost=0.2)             
pred = predict(svmmodel2, newdata=test1)
# confusion matrix
cm4 <- confusionMatrix(pred, as.factor(test1$DEATH_EVENT))
draw_confusion_matrix(cm4)
```
The two SVM models provided solid results, with the 1st being slightly superior. However, the 1st Random Forest model yielded the best results thus far.


# Decision Tree
## Model Preparation
```{r}
hf_DT <- subset(heartFailure,select=-c(id, age, platelets, serum_sodium, 
                            creatinine_phosphokinase))
str(hf_DT)
```

## Data Splitting: 1st Attempt
```{r}
set.seed(1)
train1 <- sample(1:nrow(hf_DT),209)
hf_train1 <- hf_DT[train1,]
hf_test1 <- hf_DT[-train1,]
dim(hf_train1)
dim(hf_test1)
```

## Data Splitting: 2nd Attempt
```{r}
# 2nd split
set.seed(1)
train2 <- sample(1:nrow(hf_DT),209)
hf_train2 <- hf_DT[train2,]
hf_test2 <- hf_DT[-train2,]
```

## Data Splitting: 3rd Attempt
```{r}
# 3rd split
set.seed(1)
train3 <- sample(1:nrow(hf_DT),209)
hf_train3 <- hf_DT[train3,]
hf_test3 <- hf_DT[-train3,]
```

## Data Splitting: 4th Attempt
```{r}
# 4th split
set.seed(1)
train4 <- sample(1:nrow(hf_DT),209)
hf_train4 <- hf_DT[train4,]
hf_test4 <- hf_DT[-train4,]
```

##Tree Models
```{r}
#tree models
tree_model1 <- rpart(DEATH_EVENT~.,data=hf_train1)
tree_model2 <- rpart(DEATH_EVENT~.,data=hf_train2)
tree_model3 <- rpart(DEATH_EVENT~.,data=hf_train3)
tree_model4 <- rpart(DEATH_EVENT~.,data=hf_train4)
rpart.plot(tree_model1)
rpart.plot(tree_model2)
rpart.plot(tree_model3)
rpart.plot(tree_model4)
```

## Model Fitting with Complexity Parameter vs. Cross-Validated Error
```{r}
cfit1 <- rpart(DEATH_EVENT~.,data=hf_train1, method='class')
cfit2 <- rpart(DEATH_EVENT~.,data=hf_train2, method='class')
cfit3 <- rpart(DEATH_EVENT~.,data=hf_train3, method='class')
cfit4 <- rpart(DEATH_EVENT~.,data=hf_train4, method='class')

model <- rpart(DEATH_EVENT~.,data=hf_train1, minsplit = 10, cp = 0.01)
str(model$cptable)

plotcp(model)
```

## Optimal Complexity Parameter Value
```{r}
optimal_cp <- model$cptable[which.min(model$cptable[, "xerror"]), "CP"]
str(optimal_cp)
print(optimal_cp)
plotcp(tree_model1)
plotcp(tree_model2)
plotcp(tree_model3)
plotcp(tree_model4)
```

## Final Decision Tree Model
```{r}
final_model <- rpart(DEATH_EVENT~., data = hf_train1, minsplit = 10, cp = optimal_cp)
rpart.plot(final_model)
```


# Association Rule Mining
## Model Preparation
```{r}
# Replace binary values with generic naming schemes.
binaryData["anaemia"][binaryData["anaemia"] == "0"] <- "not anaemic"
binaryData["anaemia"][binaryData["anaemia"] == "1"] <- "anaemic"
binaryData["diabetes"][binaryData["diabetes"] == "0"] <- "not diabetic"
binaryData["diabetes"][binaryData["diabetes"] == "1"] <- "diabetic"
binaryData["high_blood_pressure"][binaryData["high_blood_pressure"] == "0"] <- "stable blood pressure"
binaryData["high_blood_pressure"][binaryData["high_blood_pressure"] == "1"] <- "high blood pressure"
binaryData["sex"][binaryData["sex"] == "0"] <- "female"
binaryData["sex"][binaryData["sex"] == "1"] <- "male"
binaryData["smoking"][binaryData["smoking"] == "0"] <- "non-smoker"
binaryData["smoking"][binaryData["smoking"] == "1"] <- "smoker"
binaryData["DEATH_EVENT"][binaryData["DEATH_EVENT"] == "0"] <- "survived"
binaryData["DEATH_EVENT"][binaryData["DEATH_EVENT"] == "1"] <- "died"
head(binaryData,299) %>%
  DT::datatable()
view(binaryData)
```

## 1st Attempt
```{r}
rules <- apriori(binaryData, parameter = list(supp = 0.001, conf = 0.9))
summary(rules)
```
Using the Apriori algorithm, we want about 20-30 generated rules. This 1st attempt generated 202 rules, so parameter adjustment will be made.


## 2nd Attempt
```{r}
rules2 <- apriori(binaryData, parameter = list(supp = 0.1, conf = 0.9))
summary(rules2)
```
After increasing the support parameter, 32 rules were generated, which is slightly above our threshold. So, parameters will be adjusted again.


## 3rd Attempt
```{r}
rules3 <- apriori(binaryData, parameter = list(supp = 0.13, conf = 0.9))
summary(rules3)
inspect(rules3)
inspect(head(rules3))
```
After increasing the support parameter once more, 20 rules were generated. Since it is within the threshold of 20-30, no further parameter adjustments will be needed.


## Top 5 Rules (Confidence)
```{r}
topRules <- sort(rules3, by="confidence", decreasing = TRUE)
inspect(topRules[1:5])
```

## Top 5 Rules (Lift)
```{r}
topRules2 <- sort(rules3, by="lift", decreasing = TRUE)
inspect(topRules2[1:5])
```

## Top 5 Rules (Support)
```{r}
topRules3 <- sort(rules3, by="support", decreasing = TRUE)
inspect(topRules3[1:5])
```

## Setting items to RHS (Confidence)
### RHS for Death Event
```{r}
RHS <- apriori(binaryData, parameter = list(supp = 0.004, conf = 0.9),
               appearance = list(default = "lhs", 
                                 rhs = c("DEATH_EVENT=survived",
                                         "DEATH_EVENT=died")),
               control = list(verbose = F))
RHS <- sort(RHS, by = "support", decreasing = TRUE)
inspect(head(RHS[1:4]))
```

### RHS for Smoking
```{r}
RHS_2 <- apriori(binaryData, parameter = list(supp = 0.13, conf = 0.9),
                 appearance = list(default = "lhs", rhs = c("smoking=non-smoker", "smoking=smoker")),
                 control = list(verbose = F))
RHS_2 <- sort(RHS_2, by = "confidence", decreasing = TRUE)
inspect(head(RHS_2))
```

### RHS for Sex
```{r}
RHS_3 <- apriori(binaryData, parameter = list(supp = 0.13, conf = 0.9),
                 appearance = list(default = "lhs", rhs = c("sex=female", "sex=male")),control = list(verbose = F))
RHS_3 <- sort(RHS_3, by = "confidence", decreasing = TRUE)
inspect(head(RHS_3))
```

## HTMLwidget
```{r}
plot(topRules, method = "graph", engine = "htmlwidget", shading = "confidence")
plot(RHS, method = "graph", engine = "htmlwidget", shading = "confidence")
```
After plotting the top 5 rules with an HTMLwidget, it appeared that death event, indicating survival in this scenario, is pointing towards a dark red circle, which represents a strong rule 4. Rule 4 points towards patients being non-smokers, implying that being a non-smoker can lead to patient survival.

When plotting the RHS rules, it appeared that diabetic, smoker, and female are pointing towards Rule 3, which then points to death event = died.


## Plotting Parallel Coordinates
```{r}
plot(topRules, method = "paracoord")
plot(RHS, method="paracoord")
```




